fGFPCA dynamic prediction

### 2023-02-24 f-GFPCA.Rmd
prediction performance evaluation

library(here)
library(tidyverse)
library(ggplot2)
theme_set(theme_minimal())
library(fcr)
library(refund)
library(lme4)
library(ggpubr)
library(gridExtra)
set.seed(1025)
```

@@

#### Figure

@@

```{r, laplace_ped, eval=FALSE}
# clean MLE out-of-sample prediction results
out_pred_clean_laplace <- function(subjID=1, mid=mid){
  # Prediction
  pred_t195 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=195) %>% select(-eta_i))
  pred_t395 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=395) %>% select(-eta_i))
  pred_t595 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=595) %>% select(-eta_i))

@@

```{r, fig.width=8, fig.height=4}
# compare point prediction with in-sample
# subject 1
grid.arrange(
  out_pred_subj1 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 out-of-sample MLE"),

  in_pred_subj1 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 in-sample"),
  nrow=1)

# subject 2
grid.arrange(
  out_pred_subj2 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 out-of-sample MLE"),

  in_pred_subj2 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 in-sample"),
  nrow=1)

# subject 3
grid.arrange(
  out_pred_subj3 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 out-of-sample MLE"),

  in_pred_subj3 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 in-sample"),
  nrow=1)

# subject 4
grid.arrange(
  out_pred_subj4 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 out-of-sample MLE"),

  in_pred_subj4 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 in-sample"),
  nrow=1)

@@

#### Predictive performance measures

- Modeled after dynamic prediction paper (Leroux 2017), focusing on prediction of individual tracks
- Integrated error: for latent process

$$\frac{1}{N}\sum_{i=1}^N\sum_s(\hat{\eta}_{is}-\eta_{is})^2$$
where s covers only unobserved tracks

- Integrated AUC: for binary outcome
- Also, compare with a existing method (e.g, AMM, FRI)
- Perhaps, also computation time?

<span style="color:red;">The problem with measures above is that predictions are made on a different grid than observations. I can think of two potential ways to address this issue: 1) only evaluate at bin midpoints; 2) use interpolation to extend latend function from binned grid to the original grid (in GFPCA paper this is done through projection. ) </span>

```{r}
df_true_eta %>% filter(id==1)
eta_hat_mat <- matrix(NA, ncol = )

out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)$eta_pred
```

### In-/Out-of-sample PC scores

@@

# scores
in_score <- matrix(NA, nrow=N, ncol=4)
out_score_MLE <- matrix(NA, nrow=N, ncol=4)
out_score_laplace <- matrix(NA, nrow=N, ncol=4)

for(i in 1:N){
  in_score[i, ] <- in_samp_dyn_pred(fpca_fit=fpca_fit, 
                   df_part=df_pred_unique %>% filter(id==i & bin<=595))$score
  out_score_MLE[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
                                  select(-eta_i),
                                  fpca_fit=fpca_fit)$score_out
  out_score_laplace[i,] <- out_pred_laplace(fpca_fit=fpca_fit, 
                                            df_new = df %>% filter(id==i & sind_inx<=595) %>% select(-eta_i))$score_out
}

# bias
in_bias <- in_score-xi
out_bias_MLE <- out_score_MLE-xi
out_bias_laplace <- out_score_laplace-xi
```

```{r, fig.height=4, fig.width=8}
grid.arrange(
  out_bias_MLE %>% data.frame() %>%
    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
    ggplot()+
    geom_boxplot(aes(x=PC, y=Bias))+
    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
    labs(title = "Out-of-sample MLE"),

  out_bias_laplace %>% data.frame() %>%
    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
    ggplot()+
    geom_boxplot(aes(x=PC, y=Bias))+
    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
    labs(title = "Out-of-sample laplace"),

>>>>

library(here)
library(tidyverse)
library(ggplot2)
theme_set(theme_minimal())
library(fcr)
library(refund)
library(lme4)
library(ggpubr)
library(gridExtra)
library(kableExtra)
set.seed(1025)
```

@@

@@

```{r, laplace_pred, eval=FALSE}
# clean MLE out-of-sample prediction results
out_pred_clean_laplace <- function(subjID=1, mid=mid){
  # Prediction
  pred_t195 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=195) %>% select(-eta_i))
  pred_t395 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=395) %>% select(-eta_i))
  pred_t595 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=595) %>% select(-eta_i))

@@

```{r, fig.width=8, fig.height=4}
# compare point prediction with in-sample
# subject 1
grid.arrange(
  out_pred_subj1 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 out-of-sample prediction"),

  # out_pred_lap_subj1 %>% 
  #   group_by("type") %>%
  #   ggplot()+
  #   geom_line(aes(x=mid, y=true), col = "black")+
  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
  #             show.legend = FALSE)+
  #   labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),

  in_pred_subj1 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 in-sample estimation"),
  nrow=1)

# subject 2
grid.arrange(
  out_pred_subj2 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 out-of-sample prediction"),

  # out_pred_lap_subj2 %>% 
  #   group_by("type") %>%
  #   ggplot()+
  #   geom_line(aes(x=mid, y=true), col = "black")+
  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
  #             show.legend = FALSE)+
  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),

  in_pred_subj2 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 in-sample estimation"),
  nrow=1)

# subject 3
grid.arrange(
  out_pred_subj3 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 out-of-sample prediction"),

  # out_pred_lap_subj3 %>% 
    # group_by("type") %>%
    # ggplot()+
    # geom_line(aes(x=mid, y=true), col = "black")+
    # geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
    #           show.legend = FALSE)+
    # labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),

  in_pred_subj3 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 in-sample estimation"),
  nrow=1)

# subject 4
grid.arrange(
  out_pred_subj4 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 out-of-sample prediction"),

  # out_pred_lap_subj4 %>% 
  #   group_by("type") %>%
  #   ggplot()+
  #   geom_line(aes(x=mid, y=true), col = "black")+
  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
  #             show.legend = FALSE)+
  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
  #

  in_pred_subj4 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 in-sample estimation"),
  nrow=1)

@@

## Predictive performance measures

This section evaluates the performance of out-of-sample prediction of individual tracks. Out reference method is Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive), with a fixed effect of time and subject-level random intercept and slope:

$$g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$ 
We use two metrics for evaluation:

1. Integrated squared error.

- Evaluation of the latent Gaussian function
- For a single subject, only evaluated at unobserved bin midpoints

$$\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_{is}-\eta_{is})^2$$

2. Average AUC

- Evaluation of the binary outcome function
- AUC for a single subject is calculated over the unobserved time points on the original un-binned grid
- Predicted probability between bin midpoints are estimated with interpolation
- Please note that the grid ends at the maximum bin midpoint, instead of the maximum observed time, due to how interpolation works

3. Computation time

This is a bit more complicated. The full compuation time consists of two parts: 
- Time spent on model fitting
- Time spent on prediction **given** the fitted FPCA or adaptive GLMM models.

```{r, results='hide', message=FALSE}
source(here("Code/PredEval.R"))
```

```{r}
# table displaying predictive performance
tb_pred_eval <- tibble(
  c("195", "395", "595"),
  ise_fgfpca, ise_adglmm, auc_fgfpca, auc_adglmm, comp_t_fgfpca, comp_t_adglmm)
colnames(tb_pred_eval) <- 
  c("Maximum observed time", rep(c("fGFPCA", "GLMMadaptive"), 3))
```

```{r, results='asis'}
tb_pred_eval %>% kable(digits = 3, align = "lcccccc") %>%
  kable_styling(full_width = F) %>%
  add_header_above(c(" "=1, "ISE"=2, "AUC "=2, "Prediction time"=2))
```

## In-/Out-of-sample PC scores

@@

# scores
in_score <- matrix(NA, nrow=N, ncol=4)
out_score_MLE <- matrix(NA, nrow=N, ncol=4)
# out_score_laplace <- matrix(NA, nrow=N, ncol=4)

for(i in 1:N){
  in_score[i, ] <- in_samp_dyn_pred(fpca_fit=fpca_fit, 
                   df_part=df_pred_unique %>% filter(id==i & bin<=595))$score
  out_score_MLE[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
                                  select(-eta_i),
                                  fpca_fit=fpca_fit)$score_out
  # out_score_laplace[i,] <- out_pred_laplace(fpca_fit=fpca_fit, 
  #                                           df_new = df %>% filter(id==i & sind_inx<=595) %>% select(-eta_i))$score_out
}

# bias
in_bias <- in_score-xi
out_bias_MLE <- out_score_MLE-xi
# out_bias_laplace <- out_score_laplace-xi
```

```{r, fig.height=4, fig.width=6}
grid.arrange(
  out_bias_MLE %>% data.frame() %>%
    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
    ggplot()+
    geom_boxplot(aes(x=PC, y=Bias))+
    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
    labs(title = "Out-of-sample MLE"),

  # out_bias_laplace %>% data.frame() %>%
  #   pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
  #   ggplot()+
  #   geom_boxplot(aes(x=PC, y=Bias))+
  #   stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
  #   labs(title = "Out-of-sample laplace"),
